LCS Tool : A Computational Platform for Lagrangian Coherent Structures

### 2014-08-23 main.tex
Update main.tex for lualatex and commath
Incorporate revisions from arXiv e-print
http://arxiv.org/abs/1406.3527v1

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage[uniquename=init,firstinits,backend=bibtex,sorting=none]{biblatex}
\usepackage{calc}
\usepackage{color}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage[numbered,framed]{mcode}
\usepackage[space]{grffile}
%% hyperref manual advises loading this package last
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,breaklinks=true}

@@

\newcommand{\reals}{\mathbf R}

\title{LCS Tool — An Algorithmic Introduction to Lagrangian Coherent 
Structures}

@@

\abstract{
We give an applied introduction to Lagrangian coherent structures (LCSs) using a computational engine we have programmed called LCS Tool. First, we review LCS theory that is based on variational calculus; this introduces elliptic and hyperbolic variational LCSs. Then we turn our attention to algorithms for computations conceived in accordance with the variational LCS theory. These algorithms are the foundation of LCS Tool. We conclude by explaining, with demonstration scripts, how to use LCS Tool to analyze three flow examples: a double gyre, a Bickley jet and an ocean dataset.
}

@@

LCSs are used in fluid dynamics to capture essential flow features and thereby simplify the description of complex fluid flows. Here we focus on analyzing two dimensional finite time turbulent flows. LCSs identify transport barriers, repelling and attracting structures and vortex boundaries. Application areas for LCSs include oceanic and atmospheric flows, biological applications\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor} and mechanical systems\parencite{hadjighasem13:_detec_kam}.

The finite time Lyapunov exponent (FTLE) is one widely-used mathematical definition for LCSs. Despite its widespread use, examples where the FTLE fails to identify LCSs correctly are known\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. To remedy these problems, variational calculus has been applied to material curve properties in the context of fluid mechanics. Thus we obtain a rigorous approach to LCS detection\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,farazmand13:_attrac_lagran,haller13:_coher_lagran}. This article presents an algorithmic introduction to variational LCSs supported by a computational engine called LCS Tool\footnote{LCS Tool is available at: \url{https://github.com/jeixav/LCS-Tool}.}.

LCS Tool is a library of MATLAB functions that perform computations to identify LCSs in two dimensional finite time flows. LCS Tool aims to help scientists perform variational LCS calculations. The examples we present have been programmed into demonstration scripts. We encourage readers to download and execute these scripts after reading this article.

Other LCS software has been developed before; the list includes:
\begin{itemize}
\item \emph{MANGEN}\parencite{lekien03:_time}, calculates stability manifolds adaptively in two dimensional finite time velocity fields. Includes a graphical user interface and uses MPI for parallel calculations.
\item \emph{LCS MATLAB Kit}\parencite{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\parencite{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic and dataset defined velocity definitions.
\item \emph{FlowVC}\parencite{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
\end{itemize}
These packages are based on the FTLE; the focus of LCS Tool and this article is to encourage and facilitate variational LCS analysis.

@@

The fluid flows considered are defined in two dimensions by a velocity field:
\begin{align*}
\dot{\boldsymbol x} = \boldsymbol v(\boldsymbol x,t), && \boldsymbol x \in U \subset \reals^2, && t \in [t_-,t_+].
\end{align*}
Integration yields the flow map, defined between an initial time $t_0$ and a final time $t$ within $[t_-,t_+]$:
\[
F_{t_0}^t(\boldsymbol x_0) \equiv \boldsymbol x(t,t_0,\boldsymbol x_0).
\]
The time interval used to detect LCSs is arbitrary; it can be the complete interval for which the flow is defined, or data is available, or it can be a sub-interval.

The right Cauchy-Green strain tensor measures Lagrangian strain in the velocity field and is defined:
\[
C_{t_0}^t(\boldsymbol x_0) = \left[D F_{t_0}^t(\boldsymbol x_0)\right]^T D F_{t_0}^t(\boldsymbol x_0).
\]
The eigenvalues of the Cauchy-Green strain tensor are denoted by $\lambda_{1,2}$ and the eigenvectors by $\boldsymbol \xi_{1,2}$. The eigenvalues satisfy $0 < \lambda_1 \leq \lambda_2$ and the eigenvectors satisfy $\boldsymbol \xi_1 \perp \boldsymbol \xi_2$.

\subsection{Elliptic LCS}
We find coherent Lagrangian vortices as closed orbits of one of the two vector fields
\begin{equation}
\boldsymbol \eta^{\lambda}_\pm = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \boldsymbol \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 - \lambda_1}} \boldsymbol \xi_2,
\label{eq:eta}
\end{equation}
which are a function of the eigenvalues and eigenvectors of $C_{t_0}^t(\boldsymbol x_0)$. Curves $\gamma$ that are trajectories of $\boldsymbol \eta^{\lambda}_\pm$ can be parametrized as $\boldsymbol r(s)$ with their arc length $s \in [0,\sigma]$. They are given by integrating
\begin{equation}
\boldsymbol r' \equiv \frac{d \boldsymbol r}{ds} = \boldsymbol \eta^{\lambda}_\pm(\boldsymbol r).
\label{eq:etafields}
\end{equation}

$\lambda$ is a free parameter close to unity that indicates the uniform stretching of the curve $\gamma$, i.e., each tangent vector $\boldsymbol r'(s)$ with length $l_{t_0}$ stretches by a factor $\lambda$ to length $l_{t}$ under the flow map:
\begin{gather*}
l_{t_0}(s) = \sqrt{\langle \boldsymbol r'(s), \boldsymbol r'(s) \rangle},\\
l_t(s) = \sqrt{\langle \boldsymbol r'(s), C_{t_0}^t(\boldsymbol r(s)) \boldsymbol r'(s) \rangle},\\
l_t(s) = \lambda\, l_{t_0}(s).
\end{gather*}
For $\lambda=1$, $\boldsymbol \eta^{\lambda}_\pm$ points in the direction of maximal Lagrangian shear and we obtain uniformly non-stretching curves over the finite time interval $[t_0, t]$. In summary, the boundaries of coherent Lagrangian vortices are closed curves that uniformly stretch by a factor $\lambda$ when advected by the flow. We refer to these curves as lambda-lines. The outermost lambda-line is the physically observable boundary of an eddy, generating the tracer pattern of an isolated rotating fluid region.

Note that the equation for $\boldsymbol \eta^{\lambda}_\pm$ follows from the variational principle that the first variation of the averaged Lagrangian strain along the sought extremal closed curve vanishes (see \parencite{haller13:_coher_lagran} for full derivations):
\begin{eqnarray*}
Q(\gamma) &=& \frac{1}{\sigma} \int_0^\sigma \frac{l_t(s)}{l_{t_0}(s)}\text{d}s\\
\delta Q(\gamma) &=& 0.
\end{eqnarray*}
This coherence principle means the sought curve $\gamma$, the coherent eddy boundary, is a stationary point of the averaged Lagrangian strain functional and therefore nearby fluid elements cannot separate from $\gamma$.

\subsection{Hyperbolic LCS}
Hyperbolic LCSs are trajectories of the eigenvector fields $\boldsymbol \xi_{1,2}$. Repelling hyperbolic LCSs, denoted strainlines, are obtained as solutions of the differential equation
\begin{equation}
\boldsymbol r' = \boldsymbol \xi_1(\boldsymbol r),
\label{eq:strainline}
\end{equation}
and attracting hyperbolic LCSs, denoted stretchlines, are obtained as a solution of the differential equation
\[
\boldsymbol r' = \boldsymbol \xi_2(\boldsymbol r).
\]
A normal perturbation a strainline segment grows under the flow map by a factor $\lambda_2$, i.e., fluid elements are repelled at that rate in the normal direction. Equivalently, a normal perturbation a stretchline segment shrinks by a factor $\lambda_1$ and fluid elements are attracted.

Similarly to elliptic LCS, hyperbolic LCS are also based on a variational principle, in this case for the Lagrangian shear
\[
p(s) = \frac{\langle \boldsymbol r'(s), D(\boldsymbol r(s)) \boldsymbol r'(s)\rangle}{\sqrt{\langle \boldsymbol r'(s), C(\boldsymbol r(s)) \boldsymbol r'(s)\rangle \langle \boldsymbol r'(s), \boldsymbol r'(s)\rangle}}
\]
where
\[
D(\boldsymbol x_0) = \frac12[C(\boldsymbol x_0) \Omega - \Omega C(\boldsymbol x_0)], \quad \Omega = \begin{pmatrix}0&-1\\1&0\end{pmatrix}.
\]
The first variation of the average Lagrangian shear along a hyperbolic LCS vanishes
\[
P(\gamma) = \frac{1}{\sigma} \int_0^\sigma p(s) \text{d}s,
\]
\[
\delta P(\gamma) = 0.
\]
They are one type of shearless LCS. For details of the derivation of hyperbolic LCS, see \cite{farazmand13:_shearless}.

@@

\section{Methods}

This section presents the numerical methods used to compute LCSs and explains them step-wise. In other words, the algorithms that constitute LCS Tool are presented.

\subsection{The Cauchy-Green strain tensor}

To obtain Lagrangian coherent structures the Cauchy-Green strain tensor must be calculated. More specifically, the eigenvalues and eigenvectors of the Cauchy-Green strain tensor are needed. The function associated with this calculation in LCS Tool is \lstinline!eig_cgStrain!. The main steps of this function are enumerated in Table~\ref{t:Cauchy-Green algorithm} and are described in detail below. Table~\ref{t:eig_cgStrain syntax} summarizes the syntax of \lstinline!eig_cgStrain!.

\begin{table}
\begin{enumerate}
\item Setup a Cartesian grid and auxiliary grid in the (rectangular) domain of the flow.
\item Integrate the flow velocity equations at each grid point and auxiliary grid point. This yields the flow map, $F_{t_0}^t(\boldsymbol x_0)$.
\item Use finite differences to approximate the derivative of the flow map, $D F_{t_0}^t(\boldsymbol x_0)$.
\item Compute the Cauchy-Green strain tensor, $C_{t_0}^t = \left(D F_{t_0}^t\right)^T D F_{t_0}^t$ and its eigenvalues, $\lambda_{1,2}$ and eigenvectors, $\xi_{1,2}$.
\end{enumerate}
\caption{Algorithm to calculate the Cauchy-Green strain tensor eigenvalues and eigenvectors}
\label{t:Cauchy-Green algorithm}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\\
\hline
\lstinline!derivative! & function handle for flow velocity equations\\
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\\
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default is $10^{-2}$.\\
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default is \lstinline!true!.\\
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default is \lstinline!false!.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\\
\hline \hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

We setup a Cartesian grid according to user-defined resolutions. We can find the optimal resolution by an iterative procedure in which we start with a low resolution and double it until convergence of LCSs is observed visually. If the chosen domain only comprises a few expected LCSs, e.g., one vortex, a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, for larger domains, a higher resolution must be chosen. Furthermore an auxiliary grid is setup. This is illustrated in Figure~\ref{f:main and auxiliary grids}. The improvements the auxiliary grid yields compared to using only the main grid were reported in \textcite{farazmand12:_comput_lagran}. Experience indicates that optimal results are obtained when the auxiliary grid spacing is 1\% to 10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of main grid (filled circles) and auxiliary grid (empty circles) used to compute the Cauchy-Green strain. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

The function \lstinline!eig_cgStrain! provides the option to calculate eigenvectors from the auxiliary grid, but eigenvalues from the main grid. We have found that for flows defined analytically the eigenvalues should be calculated from the main grid whereas for the flows defined by datasets, using the auxiliary grid gives better results.

As stated, typically we set the main grid resolution of the Cauchy-Green strain tensor to around 500 grid points. For a square domain, this implies the velocity equations must be integrated, after including 4 auxiliary grid points around every main grid point, at 1.25 million grid points. The natural way to perform these integration calculations would be to solve a two dimensional first-order ODE system at every grid point and iterate over every grid point in a for-loop. With 1.25 million grid points this calculation can take an excessively long time. To alleviate this problem, we solve the flow equations in a vector form where the equations at all grid points are combined into a single system of equations, even though the system is composed of independent blocks of two dimensional first-order ODE systems. MATLAB's \lstinline!ode45! function is used to perform the integration and the flow integration typically takes five to ten minutes with default error tolerances. The drawback of vector form integration is that memory requirements can become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two dimensional form.

The Cauchy-Green strain tensor can have singularities, i.e., $C_{t_0}^t(\boldsymbol x_0) = I$. Such points are isolated in general\parencite{delmarcelle94}, therefore no special accommodations are made in computations for singularities of the Cauchy-Green strain tensor.

@@

Flows that are incompressible, meaning $D \cdot \boldsymbol v = 0$, satisfy the relation $\lambda_1(\boldsymbol x_0) \lambda_2(\boldsymbol x_0) = 1, \forall \: \boldsymbol x_0 \: \in \: U$\parencite{arnold78:_mathem}. This gives the possibility of imposing incompressibility when calculating the eigenvalues of the Cauchy-Green strain tensor by first calculating $\lambda_2$ and then setting $\lambda_1(\boldsymbol x_0) = 1/\lambda_2(\boldsymbol x_0)$. This procedure should be applied carefully since by changing $\lambda_1$ but not $\boldsymbol \xi_1$, the error in satisfying the eigenvalue equation for the Cauchy-Green strain tensor may increase.

At some grid points it may occur that $\lambda_2 < 1$ due to numerical integration errors of the flow velocity equations. By setting integration tolerances to smaller values the number of grid points with $\lambda_2 < 1$ can be reduced. Nonetheless, some grid points can be so close to singularities of the Cauchy-Green strain tensor that satisfying $\lambda_2 \geq 1$ everywhere could be excessively computationally costly. Therefore the function \lstinline!eig_cgStrain! records the number of points at which incompressibility cannot be imposed and this measure can guide setting appropriate integration tolerances.

>>>>

\usepackage{amsmath}
\usepackage[uniquename=init,firstinits,backend=bibtex,sorting=none]{biblatex}
\usepackage{calc}
\usepackage{color}
\usepackage{commath}
\usepackage{fontspec}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage[numbered,framed]{mcode}
\usepackage[space]{grffile}
\usepackage[hidelinks]{hyperref}

\hypersetup{breaklinks=true}

@@

\title{LCS Tool : A Computational Platform for Lagrangian Coherent Structures}

@@

\abstract{
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool. LCSs are most repelling, attracting and shearing material lines that form the centerpieces of observed tracer patterns in two-dimensional unsteady dynamical systems. LCS Tool implements the latest geodesic theory of LCSs for two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations. After a review of the underlying theory, we explain the steps and numerical methods used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples.
}

\tableofcontents

@@

Lagrangian Coherent Structures (LCSs) are evolving organizing centers of trajectory patterns in non-autonomous dynamical systems\parencite{haller00:_lagran,peacock13:_lagran,Haller2014}. Applications of LCSs include oceanic and atmospheric flows\parencite{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\parencite{gawlik09:_lagran}, crowd dynamics\parencite{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\parencite{hadjighasem13:_detec_kam}.

\textcite{haller01:_distin} proposed that ridges of the Finite-Time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling- and attracting-type) LCSs. A number of examples indeed support this principle\parencite{peacock13:_lagran}. Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \parencite{beron-vera10:_invar_lagran}).

More recent work has focused on an exact mathematical formulation of the properties defining LCSs\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,haller13:_coher_lagran,blazevski14:_hyper_ellip_trans_barrier_three}. In two-dimensional flows, hyperbolic and parabolic LCSs turn out to be stationary curves of the averaged material shear\parencite{farazmand14:_shearless}, whereas elliptic LCSs are stationary curves of the averaged
strain\parencite{haller13:_coher_lagran}. These variational formulations lead to explicit solutions for LCSs as null-geodesics of appropriate Lorentzian metrics.

Here we present a simple algorithmic introduction to geodesic LCS detection without details on the the underlying mathematics. We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is publicly available at: \emph{\url{https://github.com/jeixav/LCS-Tool}}.}. This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present here have been programmed into demonstration scripts.

The list of publicly available software for LCS detection further
includes the following packages:
\begin{itemize}
\item \emph{MANGEN}\parencite{lekien03:_time} calculates FTLE and advects material curves in two dimensional velocity fields. It also includes a graphical user interface and uses MPI for parallel calculations.
\item \emph{LCS MATLAB Kit}\parencite{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\parencite{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic  and dataset based velocity definitions.
\item \emph{FlowVC}\parencite{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
\end{itemize}
These packages primarily target the automated generation of FTLE plots to aid the visual assessment of hyperbolic LCSs in the flow. The purpose of LCS Tool and this article is to provide an all-purpose numerical engine for the geodesic extraction of LCSs as parametrized material curves in unsteady two-dimensional flows, and extend the scope of such extraction to elliptic LCSs.

@@

We consider two-dimensional, finite-time, unsteady velocity fields of the form
\begin{align}
\od{x}{t} = v(x,t), && x \in U \subset \mathbf R^2, && t \in [t_-,t_+].
\label{eq:v(x,t)}
\end{align}
Trajectories of \eqref{eq:v(x,t)} are denoted $x(t;t_0,x_0)$, with $x_0 \in U$ denoting their initial position in the open set $U$ at an initial time $t_0 \in [t_-,t_+]$. The flow map is then defined as
\[
F_{t_0}^t(x_0) \equiv x(t;t,x_0),
\]
mapping initial positions to current positions at time $t$. The time interval $[t_-,t_+]$ is part of the definition of the finite-time dynamical system \eqref{eq:v(x,t)}. This interval may be a time scale of interest or the maximum interval over which velocity data is available from simulations or observations.

The right Cauchy-Green strain tensor associated with the flow map is defined as
\begin{equation}
C_{t_0}^t(x_0) = \left[\nabla F_{t_0}^t(x_0)\right]^T \nabla F_{t_0}^t(x_0),
\label{eq:CG}
\end{equation}
measuring Lagrangian strain in the velocity field. This tensor is symmetric and positive definite\parencite{truesdell04}. We label the eigenvalues and eigenvectors of $C_{t_0}^t(\boldsymbol x_0)$ as follows:
\[
C_{t_0}^t \xi_i = \lambda_i \xi_i, \qquad 0 < \lambda_1 \leq \lambda_2, \quad i = 1,2;
\]
\begin{equation}
\left|\xi_i\right| = 1, \quad \xi_2 = \Omega \xi_1, \quad \Omega = \left(
\begin{array}{rr}
0 & -1\\
1 & 0
\end{array}
\right).
\label{eq:CG_invariants}
\end{equation}

\subsection{Elliptic LCSs}
\label{sub:Elliptic-LCSs}

We seek positions of closed material lines at time $t_0$ that prevail as coherent Lagrangian vortex boundaries (or \emph{elliptic LCSs}) over a time interval $[t_0,t]\subset[t_-,t_+]$. \textcite{haller13:_coher_lagran} argue that such initial material line positions are closed stationary curves of the averaged strain functional
\[
Q(\gamma) = \frac{1}{\sigma} \int_0^{\sigma} \frac{\sqrt{\langle r'(s),C_{t_0}^t(r(s))r'(s)\rangle}}{\sqrt{\langle r'(s),r'(s)\rangle}} \dif s,
\]
obtained by averaging the tangential strain arising over $[t_0,t]$ along closed material lines parametrized as $r(s)$ with $s \in [0,\sigma].$
Solutions to this variational problem turn out to be closed orbits of one of two parametrized vector-field families
\begin{equation}
\eta_\pm^\lambda = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 - \lambda_1}}\xi_2,
\label{eq:eta}
\end{equation}
with $\lambda > 0$ playing the role of a parameter. Such closed orbits satisfy the differential equation
\begin{equation}
r' = \eta_\pm^\lambda(r),
\label{eq:etafields}
\end{equation}
which coincide with null-geodesics of the Lorentzian metric family $e_\lambda(u,v) = \left\langle u,\left[D_{t_0}^t(r) - \lambda^2 I\right] v \right\rangle$. For this reason, we refer to the computation of elliptic LCSs as limit cycles of \eqref{eq:etafields} as \emph{geodesic detection of elliptic LCSs}.

Any orbit of \eqref{eq:etafields} turns out to stretch uniformly under the flow map $F_{t_0}^t$. Specifically, any subset of an orbit of \eqref{eq:etafields} increases its arclength precisely by a factor of $\lambda$. For this reason, we refer to trajectories of \eqref{eq:etafields} as $\lambda$-lines. Following \parencite{haller13:_coher_lagran}, we call the outermost member of a closed family of $\lambda$-lines a coherent Lagrangian vortex boundary.

\subsection{Hyperbolic LCSs}
\label{sub:Hyperbolic-LCSs}

Next we consider positions of material lines at time $t_0$ that prevail as most repelling or attracting material lines (or \emph{hyperbolic LCSs}) over a time interval $[t_0,t] \subset [t_-,t_+]$. \textcite{farazmand14:_shearless} argue that hyperbolic LCSs are stationary curves of the averaged shear functional
\begin{gather*}
Q(\gamma) = \frac{1}{\sigma} \int_0^\sigma \frac{\langle r'(s),D_{t_0}^t(r(s)) r'(s)\rangle}{\sqrt{\langle r'(s),C_{t_0}^t(r(s)) r'(s)\rangle\langle r'(s),r'(s)\rangle}}\dif s,\\
D_{t_0}^t = \frac12 [C_{t_0}^t \Omega - \Omega C_{t_0}^t],
\end{gather*}
obtained by averaging the Lagrangian shear arising over $[t_0,t]$ along closed material lines parametrized as $r(s)$ with $s \in [0,\sigma]$. More precisely, hyperbolic LCSs are stationary curves of $Q(\gamma)$ with respect to fixed-endpoint perturbations. We note that parabolic LCSs (Lagrangian jet cores) are also stationary curves of $Q(\gamma)$, but under variable endpoint perturbations (cf. \textcite{farazmand14:_shearless}.)

Solutions to this variational problem turn out to be orbits of the $\xi_1$ or $\xi_2$ eigenvector field. Repelling LCSs (\emph{shrinklines}) are obtained as trajectories of the differential equation
\begin{equation}
r' = \xi_1(r),
\label{eq:strainline}
\end{equation}
and attracting LCSs (\emph{stretchlines}) are obtained as trajectories of the differential equation
\begin{equation}
r' = \xi_2(r).
\label{eq:stretchline}
\end{equation}

Shrinklines and stretchlines coincide with the null-geodesics of the Lorentzian metric $h(u,v) = \left\langle u,D_{t_0}^t(r) v \right\rangle$. For this reason, we refer to the computation of hyperbolic LCSs as strongest normally-repelling or normally-attracting orbits of \eqref{eq:etafields} as \emph{geodesic detection of hyperbolic LCSs}.

To compute the normal repulsion of shrinklines, we note that an infinitesimal
normal perturbation to a shrinkline $\gamma$ at its point $r$ grows under the flow map $F_{t_{0}}^{t}$ by a factor of $\lambda_{2}(r)$ in the direction normal to $F_{t_{0}}^{t}(\gamma).$ Similarly, small normal perturbations to a stretchline decay by a factor $\lambda_{1}(r)$ in the direction normal to the evolving stretchline.

@@

\section{Numerical methods}

Here we describe a step-by-step numerical implementation of the geodesic detection of elliptic and hyperbolic LCSs through the differential equations \eqref{eq:etafields}, \eqref{eq:strainline}, and \eqref{eq:stretchline}.

\subsection{Computing the invariants of the Cauchy-Green strain tensor}

The common first step in calculating elliptic and hyperbolic LCSs is the computation of the Cauchy-Green strain tensor field $C_{t_0}^t(x_0)$, as defined in \eqref{eq:CG}. The function performing this calculation in LCS Tool is \lstinline!eig_cgStrain!. The main steps executed by this function are enumerated in Table~\ref{t:Cauchy-Green algorithm}, while Table~\ref{t:eig_cgStrain syntax} summarizes the syntax of \lstinline!eig_cgStrain!.

\begin{table}
\begin{enumerate}
\item Define a Cartesian grid for initial conditions of trajectories. Define
an auxiliary grid for differentiating with respect to initial conditions.
\item Solve \eqref{eq:v(x,t)} starting from each grid point and auxiliary grid point over the time interval $[t_0,t]$. This gives a discrete approximation to the flow map $F_{t_0}^t(x_0)$.
\item Use finite differencing over the auxiliary grid to compute numerically the derivative of the flow map $DF_{t_0}^t(x_0)$.
\item Compute the Cauchy-Green strain tensor field $C_{t_0}^t(x_0) = \left(DF_{t_0}^t(x_0)\right)^T DF_{t_0}^t(x_{0})$, its eigenvalue field $\lambda_{1,2}(x_0)$, and eigenvector fields $\xi_{1,2}(x_0)$ over the initial condition grid.
\end{enumerate}
\caption{Algorithm to calculate the invariants of the Cauchy-Green strain tensor field.}
\label{t:Cauchy-Green algorithm}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\tabularnewline
\hline
\lstinline!derivative! & function handle for flow velocity equations\tabularnewline
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\tabularnewline
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\tabularnewline
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default is $10^{-2}$.\tabularnewline
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default is \lstinline!true!.\tabularnewline
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default is \lstinline!false!.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\tabularnewline
\hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

The Cartesian grid of initial conditions mentioned in Table~\ref{t:Cauchy-Green algorithm} is rectangular, with user-defined vertical and horizontal ranges and resolutions. The optimal resolution may be determined by a successive doubling of the initial resolution until convergence of the extracted LCSs is observed visually. If the domain of interest comprises only a few expected LCSs, e.g., one vortex, then a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, a higher resolution must be chosen.

The auxiliary grid (cf. Table~\ref{t:Cauchy-Green algorithm}) comprises four points placed symmetrically around each point of the Cartesian grid (Figure~\ref{f:main and auxiliary grids}.) These points are used to achieve increased accuracy in the finite-difference approximation
\[
\nabla F_{t_0}^t(x_0) \approx
\left(\begin{array}{cc}
\frac{x_1(t;t_0,x_0 + \delta x_1) - x_1(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_1(t;t_0,x_0+\delta x_2) - x_1(t;t_0,x_0-\delta x_2)}{2\vert\delta x_2\vert}\\
\frac{x_2(t;t_0,x_0+\delta x_1) - x^2(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_2(t;t_0,x_0 + \delta x_2) - x_2(t;t_0,x_0 - \delta x_2)}{2\vert\delta x_2\vert}
\end{array}\right)
\]
of $\nabla F_{t_0}^t$ at a point $x_0$ of the Cartesian grid. Here $\delta x_i$ is a vector of length $\vert\delta x_i\vert > 0$ that points from the Cartesian grid-point $x_0$ in the $i^{th}$ coordinate direction (Figure~\ref{f:main and auxiliary grids}). Computational improvements arising from the use of the auxiliary grid over simply using the nearest points of the main grid were reported in Ref.\,\parencite{farazmand12:_comput_lagran}. Experience suggests an auxiliary grid spacing of 1-10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=.5\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of the main grid (filled circles) and the auxiliary grid (empty circles) used in the computation of the derivative of the flow map in the definition \eqref{eq:CG} of the Cauchy--Green stain tensor. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

The function \lstinline!eig_cgStrain! of LCS Tool provides the option to calculate Cauchy--Green eigenvectors from the auxiliary grid using eigenvalues calculated from the main grid. We have found that for flows defined analytically, the eigenvalues can reliably be calculated from the main grid. For the flows defined by datasets, using the auxiliary grid for eigenvalue calculations gives better results.

As mentioned above, a typical main grid for the Cauchy-Green strain tensor comprises $500 \times 500$. This means that after the addition of 4 auxiliary grid points around each main grid points, eq. \eqref{eq:v(x,t)} must be integrated over 1.25 million initial conditions.

To avoid excessive computational times in MATLAB, we vectorize \eqref{eq:v(x,t)}, i.e., combine its right-hand-side evaluated over each initial points into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ODEs. We then use MATLAB's \lstinline!ode45! to perform trajectory integration from all grid points simultaneously. This process typically takes five to ten minutes with default error tolerances. A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

The Cauchy-Green strain tensor generically admits points singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined. This generically arises at a finite set of isolated points within the computational domain\parencite{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

Incompressible flows (i.e. those satisfying $\nabla \cdot v=0$) admit the relationship $\lambda_1(x_0) \lambda_2(x_0) = 1$ at all points of the computational domain \parencite{arnold78:_mathem}. Incompressibility can be computationally imposed by first calculating $\lambda_2(x_0)$, then setting $\lambda_1(x_0) = 1/\lambda_2(x_0)$ and calculating the strain eigenvectors $\xi_2$ from $\lambda_2$, then $\xi_1$ from the relationship in \eqref{eq:CG_invariants}. Experience shows that computing $\lambda_i$ in this order gives higher accuracy than in the reverse order \parencite{farazmand12:_comput_lagran}.

At some grid points, $\lambda_2 < 1$ may occur due to numerical integration errors. By setting the integration tolerances to smaller values, the number of such grid points is reduced. Enforcing $\lambda_2 \geq 1$ everywhere, however, can incur excessive computational cost. To this end, the function \lstinline!eig_cgStrain! records the number of points with $\lambda_2 < 1$, providing a measure for setting feasible integration tolerances.

\subsubsection{Special case: dataset velocity fields}
